Simulating High-Dimensional Multivariate Data

### 2020-12-16 010-background.Rmd
4th draft. Address WWP background part 2

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA express for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

@@

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

@@

*Correspondence among Pearson, Spearman, $\tau$ correlations*.
This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
But for **bivariate normal vectors**, however, the correspondence is well-known:

@@

These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

Note that with continuous marginals the rescaling returns $\rho_S$.
For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
And further the infinite sums must be approximately for count-valued data and potentially violating desired property SP2.

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
This situation gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
In general, the bounds are given by

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDF of random variables $X_1$ and $X_2$, respectively. 
For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

@@

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

>>>>

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% of highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

@@

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01].
Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

@@

*Correspondence among Pearson, Spearman, $\tau$ correlations*.
There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
For *bivariate normal vectors*, however, the correspondence is well-known:

@@

These features are critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

Note that with continuous marginals the rescaling returns $\rho_S$.
For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
Further the infinite sums must be approximated for count-valued data.

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied (see @Nelsen2007 and @BF17).
This gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
In general, the bounds are given by

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively. 
For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This provides a powerful, natural way to characterize joint probability structures.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ satisfying

@@

A Gaussian copula has marginal CDFs as the normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
